Extended Data

### 2022-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
new title!
closes #139
remove one reference (gets us to 50 limit)
#137

**Title:** Warmer spring temperatures in temperate deciduous forests advance the timing of tree growth but have little effect on annual woody productivity

@@

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem diameter growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual ring width was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting projections from global C cycle models[@richardson_terrestrial_2012;@keenan_net_2014], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**

@@

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022]. 
In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021].
Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

>>>>

**Title:** Warm springs alter timing but not total growth of temperate deciduous trees

@@

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) uptake[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;--><!--@jeong_phenology_2011;-->@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem diameter growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual ring width was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting projections from global C cycle models[@richardson_terrestrial_2012;@keenan_net_2014], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**

@@

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[<!--@jeong_phenology_2011;--><!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@keenan_net_2014;@finzi_carbon_2020;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022]. 
In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021].
Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
